---
title: "Piecewise_Regression_Lab"
author: "Tyler Skaggs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Piecewise_Regression_Lab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Intro2MLR)
library(purrr)
```

## Golden Segment Searching and Optimization

The Golden Segment Search (or Golden Section Search) is a method of finding a minimum or maximum of a function by repeatedly refining the search interval upon each iteration. If we were to search for a minimum of the function $f$ (this can be easily rephrased for a maximum) we first start with selecting some interval $[x_1, x_3]$ and select some value $x_2$ where $f(x_2)$ is a suspected minimum. in this interval such that if $x_2 - x_1 = a$ and $x_3 - x_2 = b$, then $\frac{b}{a+b} = \frac{a}{b}$. Then we then select a fourth point $x_4$ so that $x_3 - x_4 = a$. Then, we look at $f(x_4)$. If $f(x_4) < f(x_2)$, then $x_4$ is a new minimum and we redefine the interval as $[x_2, x_3]$. If $f(x_4) > f(x_2)$, then $x_2$ is a new minimum and we redefine the interval as $[x_1, x_4]$. We then repeat this process on our new interval. We can end this searching by reaching a sufficient error, which is just the distance of the interval being sufficiently small.

The name of this searching is called the golden segment search, which gets its name from the golden ratio found in the searching process. In this search we have the ratio of lengths being $\frac{b}{a} = \frac{a}{a+b}$ and to repeat this search we need the next iteration to have the same ratio. That is $\frac{a - b}{b} = \frac{b}{a}$. From here we get:
$$ \frac{a}{b} - 1 = \frac{b}{a} \rightarrow \left(\frac{a}{b}\right)^2 - \frac{a}{b} = 1$$
Of which the only positive value that satisfies this equation is the golden ratio.

## The Functions

### `findKnot()`
```{r}
findKnot <- function(xlist, ylist, data, doPlot = TRUE){

######################## Set Up ##########################
data <- data.frame(data)

x_vals <- (as.vector(data[ , xlist])) #List of x values
y_vals <- (as.vector(data[ , ylist])) #List of y values
  #Reading this in was weird as sometimes as.vector(data[ , ylist]) would be a
  #vector and other times it would be a list of a vector. This way was more consistent


min_x <- x_vals[which.min(x_vals)] #Max and min of the x values, useful later
max_x <- x_vals[which.max(x_vals)]

######################## Sub-Funcitons ##########################

findRsq <- function(x_k){ #Function for Purrr map
  Rsq <- vector(mode = "numeric", length = length(x_k))

  #For loop to return vector of Rsq values if a vector of values
  #is passed in, as in the curve function
  for(i in 1:length(x_k)){
    x <- (x_vals > x_k[i])*(x_vals - x_k[i])
    ylm <- lm(as.vector(y_vals) ~ as.vector(x_vals) + as.vector(x))
    Rsq[i] = (summary(ylm))$r.squared
  }
  return(Rsq)
}

findAIC <- function(x_k){ #Function for Purrr map
  AICs <- vector(mode = "numeric", length = length(x_k))

  #For loop to return vector of Rsq values if a vector of values
  #is passed in, as in the curve function
  for(i in 1:length(x_k)){
    x <- (x_vals > x_k[i])*(x_vals - x_k[i])
    ylm <- lm(as.vector(y_vals) ~ as.vector(x_vals) + as.vector(x))
    AICs[i] = AIC(ylm)
  }
  return(AICs)
}

######################## Calculations ##########################

Rsq <- map_dbl(x_vals, findRsq) #using purrr, yay!!
AICs <- map_dbl(x_vals, findAIC)

max_dr <- x_vals[which.max(Rsq)] #Best Rsquared is the maximum
min_da <- x_vals[which.min(AICs)] #Best AIC Value is the minimum value

c <- 3 * sqrt(max_x - min_x) #used for optimize around an interval of the discrete case
max_cr <- optimize(findRsq, interval = c(max_dr - c, max_dr + c), maximum = TRUE)
min_ca <- optimize(findAIC, interval = c(min_da - c, min_da + c), maximum = FALSE)


######################## PLOTTING ##########################

if(doPlot){ #only plot when asked (will plot by default)
  par(mfrow = c(2,1), mar = c(4,5,2,2))

  plot(x_vals, Rsq,
       xlab = expression(x[k]), ylab = expression(R^2),
       col = "black", pch = 20, lwd = 2.5)
  abline(v = max_dr, col = "red", lwd = 1)

  curve(findRsq, add = TRUE, lwd = 1.4, col = "black", n = 555) #n was arbitrary chosen
  abline(v = max_cr$maximum, col = "blue", lwd = 1)

  #---------------------------------------------------#

  plot(x_vals, AICs,
       xlab = expression(x[k]), ylab = expression(AIC),
       col = "black", pch = 20, lwd = 2.5)
  abline(v = min_da, col = "red", lwd = 1)

  curve(findAIC, add = TRUE, lwd = 1.4, col = "black", n = 555) #n was arbitrary chosen
  abline(v = min_ca$minimum, col = "blue", lwd = 1)
}

######################## RETURN ##########################

my_list <- list(x_krd = max_dr, x_krc = max_cr$maximum,
                x_kad = min_da, x_kac = min_ca$minimum,
                Rsqs = Rsq, AICs = AICs)
return(invisible(my_list))
}
```


### `pieceReg()`
```{r}
pieceReg <- function(xlist, ylist, data, type = "rd"){

  ######################## Set Up ##########################

  data <- data.frame(data)

  results <- findKnot(xlist, ylist, data, doPlot = FALSE)

  x_vals <- (as.vector(data[ , xlist])) #List of x values
  y_vals <- (as.vector(data[ , ylist])) #List of y values

  #Switch determines which knot to make model from
  x_k <- switch(type,

                rd = results$x_krd, #default case

                rc = results$x_krc,

                ad = results$x_kad,

                ac = results$x_kac

                )

  ######################## Sub-Functions ##########################

  ## Calculation of new model with 1 knot
  newmodel <- function(xk){
    X <- (x_vals - xk)*(x_vals > xk) #Second line for piecewise function based on knot
    df = within(data, X)
    newlm <- lm(as.vector(y_vals) ~ as.vector(x_vals) + as.vector(X))
    return(newlm)
  }

  ## Piecewise Model for graphing and calculations
  model <- function(xval, xk = x_k, beta = newbeta){
    beta[1] + beta[2] * xval + ((xval > xk) * beta[3] * (xval - xk))
  }

  ######################## Calculations ##########################

  pilm <- newmodel(x_k) #Find New model
  newbeta <- unname(coef(pilm)) #Coefficients of model

  x_rsq <- (summary(pilm))$r.squared
  x_aic <- AIC(pilm)

  y_k = model(x_k, x_k, newbeta) #find position of knot

  ######################## Plotting ##########################

  par(mfrow = c(1,1))

  plot(x_vals, y_vals, xlab = xlist, ylab = ylist,
       col = "black", pch = 20, lwd = 2)
  curve(model, add = TRUE, n = 1000, col = "red", lwd = 2.25)
  points(x_k, y_k, col = "blue", pch = 4, lwd = 2, cex = 1.25)


  ######################## Return ##########################

  my_list <- list(cord = c(x_k, y_k), r.squared = x_rsq, AIC =  x_aic)
  return(invisible(my_list))
}
```


## Examples
```{r, out.width="100%", fig.width=8, fig.height = 6, fig.align='center'}
v <- Intro2MLR::myreadxl("C:/Users/tyler/Documents/Grad School/MATH 5773/Data/Excel/")
t <- findKnot("LENGTH", "WEIGHT", data = v$DDT)
t
```

Using this function, we write the second function `pieceReg()`. Here is an example of it:
```{r, out.width="100%", fig.width=8, fig.height = 6, fig.align='center'}
t <- pieceReg("LENGTH", "WEIGHT", v$DDT)
t
```

Both of these functions are written general to any data set. That is, if we wanted to test
some other data set for a 2-piece piecewise regression, it is as simple as the following:
```{r, out.width="100%", fig.width=8, fig.height = 6, fig.align='center'}
t <- pieceReg("BHDiameter", "Height", v$SPRUCE, type = "rc")
t
```



